National demographic-disturbance relationships and local monitoring both reduce uncertainty in projections of boreal caribou population viability: tools to support monitoring decisions

### 2023-08-30 analysis/paper/paper.Rmd
chipping away at writing.

For this investigation we assume collars are deployed in January, and last for up to 6 years. The target number of collared females $n$ varies among scenarios, as does the total duration of the monitoring program (Table \@ref(tab:monitoringScenarios)). Collars lost to mortality may be replaced each year ($o=1$), or there may be a 3 year gap between deployments ($o=4$). We focus on exploring a set of scenarios in which anthropogenic disturbance, measured as the percentage of area within 500 m of disturbance [@johnson_science_2020;@eccc_scientific_2011], increases over time from a variety of starting points (Table \@ref(tab:monitoringScenarios)). Scenarios in which anthropogenic disturbance does not increase over time are included in Supplement D.

@@

@@

```{r figStatusErrors, fig.cap="The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure 1 for scenarios). Probabilities are calculated for 200 example population trajectories (see example trajectories in Figure 2). For each example population, the status assessment is correct if the the Bayesian posterior mean and the true population growth rate are both less than or greater than one. When disturbance is very high the probability of correct status assessment is high even without local monitoring. Where disturbance is moderate, increasing the number of years of monitoring and the number of collars reduces uncertainty, but there are diminishing returns on additional monitoring as sample size increases.", cache=TRUE}
knitr::include_graphics(here::here("figs/forPaper/powercmult6ay3aSf1.png"))
```

@@

>>>>

For this investigation we assume collars are deployed in January, and last for up to 6 years. The target number of collared females $n$ varies among scenarios, as does the total duration of the monitoring program (Table \@ref(tab:monitoringScenarios)). Collars lost to mortality may be replaced each year ($o=1$), or there may be a 3 year gap between deployments ($o=4$). We focus on exploring a set of scenarios in which anthropogenic disturbance, measured as the percentage of area within 500 m of disturbance [@johnson_science_2020;@eccc_scientific_2011], increases over time from a variety of starting points (Table \@ref(tab:monitoringScenarios), Figure \@ref(fig:figDisturbance)). Scenarios in which anthropogenic disturbance does not increase over time are included in Supplement D.

@@

In order to assess whether a monitoring program is likely sufficient to inform decisions, it is important to select outcome measures that are relevant to decision making [@gregory_structured_2012]. For boreal caribou and other endangered species, it is important to know whether a population is declining or not. If uncertainty about population status is low, additional investment in reducing uncertainty about the population growth rate may not be necessary or warranted. Thus, in this analysis our outcome of interest is how often the Bayesian posterior mean correctly predicts the true population status (increasing or decreasing). To calculate this outcome for each disturbance and monitoring scenario (Figure \@ref(fig:figDisturbance)) we fit the Bayesian model to 200 example population trajectories (e.g. Figure \@ref(fig:exampleTrajectories)). The probability of correct status assessment for that scenario is the percentage of these 200 examples in which the Bayesian posterior mean population growth rate and the true population growth rate are both above or below 1.

@@

```{r figStatusErrors, fig.cap="The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \\@ref(fig:figDisturbance) for scenarios). Probabilities are calculated for 200 example population trajectories (see example trajectories in Figure \\@ref(fig:exampleTrajectories)). For each example population, the status assessment is correct if the the Bayesian posterior mean and the true population growth rate are both less than or greater than one. When disturbance is very high the probability of correct status assessment is high even without local monitoring. Where disturbance is moderate, increasing the number of years of monitoring and the number of collars reduces uncertainty, but there are diminishing returns on additional monitoring as sample size increases.", cache=TRUE}
knitr::include_graphics(here::here("figs/forPaper/powercmult6ay3aSf1.png"))
```

@@

Include some commentary on addressing limitations of aspatial demographic models with spatial models. Not a panacea. It is possible (cite examples) to develop spatially explicit models of movement and adult mortality risk - though sample size is a challenge. Recruitment, which is more variable among ranges and with anthropogenic disturbance, is difficult to characterize at finer spatial scales because it depends on the conditions experienced by mothers over years as they move across the range, and on spatial variation in calf mortality risk that is not measured directly.
